Approximate Bayesian Computation in Large Scale Structure: Constraining the Galaxy-Halo Connection

### 2016-01-27 halo/text/abc.tex
text update

ss[12pt, preprint]{aastex}
\usepackage[breaklinks,colorlinks, urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{graphicx}	% For figures
\usepackage{natbib}	% For citep and citep
\usepackage{amsmath}	% for \iint
\usepackage{bm}
\usepackage[breaklinks]{hyperref}	% for blackboard bold numbers
\usepackage{hyperref}
\usepackage{algorithmic,algorithm}
\hypersetup{colorlinks}
\usepackage{color}
\usepackage{morefloats}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{ colorlinks,
linkcolor=darkblue,
filecolor=darkgreen,
urlcolor=darkred,
citecolor=darkblue}

@@

\begin{document}
% We need a better definitely shorter title...
\title{Approximate bayesian computation in large scale structure modeling : constraining the galaxy-halo connection}
\begin{abstract}
\todo{This is not a paper about hod, let's not start with describing hod. I suggest that we talk about parameter estimation in large scale structure in general and then move on to hod. Constraining hod is a just the first step in testing our methodology.} The Halo Occupation Distribution (HOD) model is a powerful prescription of how galaxies
populate dark matter halos that has had much success in reproducing large scale galaxy
clustering and understanding galaxy evolution within the context of the halo model.
Constraints on the parameters that describe the HOD model have so far been inferred using 
the standard practice of Bayesian inference with likelihood estimation that assumes a
functional Gaussian form of the likelihood.\todo{(Do we want to include analytic models for clustering statistics here? Yes, also we should say we use summary statistics that are not independent: clustering + group. Also there is no analytic model for the abundance of galaxy group richnesses)}  % I think we should keep it simple. There's no need to go into detail about everything in the abstract. This is already a pretty sizable abstract. 
Likelihood free inference methods such as the
Approximate Bayesian Computation (ABC) makes Bayesian inference possible without ever 
having to calculate the likelihood.

@@

Cosmology was revolutionized in the 1990s with the introduction of likelihoods---%
pro\-ba\-bil\-iites for the data given the theoretical model and parameter settings---%
for combining data from different surveys and performing principled inferences of
the cosmological parameters (CITE CITE).
Nowhere has this been more true than in cosmic microwave background studies,
where it is nearly possible to analytically evaluate a likelihood function that
involves no (or minimal) approximations (CITE Wandelt and others).

Fundamentally, the tractability of likelihood functions in cosmology flows from
the fact that the initial conditions are exceedingly close to Gaussian in form,
and many sources of measurement noise are also Gaussian.
Likelihood functions are easier to write down and evaluate when things are closer to Gaussian,
so they get easier as the length scales get larger and the structure formation gets
younger.
Hence the suitability of CMB data for likelihood analyses.

@@

%\todo{Talk about the observable: 2pcf, group multiplicity %function, and number density ...}

\section{Mock Observations}\label{sec:data}
\todo{The fake observations we use as "data" in our inference.}
%Plots showing 2pcf, gmf, ... of the mock data.
%How we compute the covariance matrix from mocks...
%Talk about error bars. Talk about poisson errors in group %populations ...

\section{Results}\label{sec:results}
Show the posteriors...
Show evolution of the posterior errorbars with iteration. Show epsilon. Convice that we are converging ...
Draw samples from the posterior, populate the mock catalogs with those samples, and compute the summary statistcs with them 
and demonstrate that they match the data...

\subsection{Comparison to MCMC}
\todo{Apples to apples comparison between ABC and MCMC.}
ABC-PMC using nbar and xi vs MCMC

\subsection{ABC Including GMF}
ABC-PMC results using observables beyond galaxy clustering and nbar(z) GMF

@@

%%%%%%%%%% FIGURES %%%%%%%%%%

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{xi_covariance_Mr20_Nmock500.png}
\caption{eft panel:
Correlation coefficients of the error covariance matrix for the Mock galaxy two-point correlation function signal in Fig. 1, where the covariance matrix is computed from the 500 mock catalogs}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{gmf_covariance_Mr20_Nmock500.png}
\caption{eft panel:
Correlation coefficients of the error covariance matrix for the Mock galaxy group multiplicity function signal in Fig. 2, where the covariance matrix
is computed from the 500 mock catalogs}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{tpcf_sigma.pdf}
\caption{1-$\sigma$(dark-shaded) and two-$\sigma$ ABC posterior predictions
for the two-point correlation function of galaxies. The black lines show the data-points.}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

@@

\section*{acknowledgments}

We would like to thank Mike Blanton, Boris Leidstadt, Micheal Schneider, Uros Seljak, Ross Fadely, Jeremy Tinker, Alex Malz, Patrick McDonald, Dan Foreman-Mackey, for productive and insightful discussions. 
MJV was supported by NSF grant AST-1517237. CHH was supported by James Arthur Graduate Award Grant.
DWH was supported by NSF (grants IIS-1124794 and AST-1517237), NASA (grant NNX12AI50G), and the Moore-Sloan Data Science Environment at NYU.

\bibliographystyle{yahapj}
\bibliography{ccppabc}
\end{document}
\documentclass[12pt, preprint]{aastex}
\usepackage[breaklinks,colorlinks, urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{graphicx}	% For figures
\usepackage{natbib}	% For citep and citep
\usepackage{amsmath}	% for \iint
\usepackage{bm}
\usepackage[breaklinks]{hyperref}	% for blackboard bold numbers
\usepackage{hyperref}
\usepackage{algorithmic,algorithm}
\hypersetup{colorlinks}
\usepackage{color}
\usepackage{morefloats}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{ colorlinks,
linkcolor=darkblue,
filecolor=darkgreen,
urlcolor=darkred,
citecolor=darkblue}

\begin{document}
% We need a better definitely shorter title...
\title{Approximate bayesian computation in large scale structure modeling : constraining the galaxy-halo connection}
\begin{abstract}
\todo{This is not a paper about hod, let's not start with describing hod. I suggest that we talk about parameter estimation in large scale structure in general and then move on to hod. Constraining hod is a just the first step in testing our methodology.} The Halo Occupation Distribution (HOD) model is a powerful prescription of how galaxies
populate dark matter halos that has had much success in reproducing large scale galaxy
clustering and understanding galaxy evolution within the context of the halo model.
Constraints on the parameters that describe the HOD model have so far been inferred using 
the standard practice of Bayesian inference with likelihood estimation that assumes a
functional Gaussian form of the likelihood.\todo{(Do we want to include analytic models for clustering statistics here? Yes, also we should say we use summary statistics that are not independent: clustering + group. Also there is no analytic model for the abundance of galaxy group richnesses)}  % I think we should keep it simple. There's no need to go into detail about everything in the abstract. This is already a pretty sizable abstract. 
Likelihood free inference methods such as the
Approximate Bayesian Computation (ABC) makes Bayesian inference possible without ever 
having to calculate the likelihood.

Cosmology was revolutionized in the 1990s with the introduction of likelihoods---%
pro\-ba\-bil\-iites for the data given the theoretical model and parameter settings---%
for combining data from different surveys and performing principled inferences of
the cosmological parameters (CITE CITE).
Nowhere has this been more true than in cosmic microwave background studies,
where it is nearly possible to analytically evaluate a likelihood function that
involves no (or minimal) approximations (CITE Wandelt and others).

Fundamentally, the tractability of likelihood functions in cosmology flows from
the fact that the initial conditions are exceedingly close to Gaussian in form,
and many sources of measurement noise are also Gaussian.
Likelihood functions are easier to write down and evaluate when things are closer to Gaussian,
so they get easier as the length scales get larger and the structure formation gets
younger.
Hence the suitability of CMB data for likelihood analyses.

%\todo{Talk about the observable: 2pcf, group multiplicity %function, and number density ...}

\section{Mock Observations}\label{sec:data}
\todo{The fake observations we use as "data" in our inference.}
%Plots showing 2pcf, gmf, ... of the mock data.
%How we compute the covariance matrix from mocks...
%Talk about error bars. Talk about poisson errors in group %populations ...

\section{Results}\label{sec:results}
Show the posteriors...
Show evolution of the posterior errorbars with iteration. Show epsilon. Convice that we are converging ...
Draw samples from the posterior, populate the mock catalogs with those samples, and compute the summary statistcs with them 
and demonstrate that they match the data...

\subsection{Comparison to MCMC}
\todo{Apples to apples comparison between ABC and MCMC.}
ABC-PMC using nbar and xi vs MCMC

\subsection{ABC Including GMF}
ABC-PMC results using observables beyond galaxy clustering and nbar(z) GMF

%%%%%%%%%% FIGURES %%%%%%%%%%

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{xi_covariance_Mr20_Nmock500.png}
\caption{eft panel:
Correlation coefficients of the error covariance matrix for the Mock galaxy two-point correlation function signal in Fig. 1, where the covariance matrix is computed from the 500 mock catalogs}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{gmf_covariance_Mr20_Nmock500.png}
\caption{eft panel:
Correlation coefficients of the error covariance matrix for the Mock galaxy group multiplicity function signal in Fig. 2, where the covariance matrix
is computed from the 500 mock catalogs}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{tpcf_sigma.pdf}
\caption{1-$\sigma$(dark-shaded) and two-$\sigma$ ABC posterior predictions
for the two-point correlation function of galaxies. The black lines show the data-points.}
\label{fig:2pcf-model}
\end{center}
\end{figure*}

\section*{acknowledgments}

We would like to thank Mike Blanton, Boris Leidstadt, Micheal Schneider, Uros Seljak, Ross Fadely, Jeremy Tinker, Alex Malz, Patrick McDonald, Dan Foreman-Mackey, for productive and insightful discussions. 
MJV was supported by NSF grant AST-1517237. CHH was supported by James Arthur Graduate Award Grant.
DWH was supported by NSF (grants IIS-1124794 and AST-1517237), NASA (grant NNX12AI50G), and the Moore-Sloan Data Science Environment at NYU.

>>>>

\documentclass[12pt, preprint]{aastex}
\usepackage[breaklinks,colorlinks, urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{graphicx}	% For figures
\usepackage{natbib}	% For citep and citep
\usepackage{amsmath}	% for \iint
\usepackage{bm}
\usepackage[breaklinks]{hyperref}	% for blackboard bold numbers
\usepackage{hyperref}
\usepackage{algorithmic,algorithm}
\hypersetup{colorlinks}
\usepackage{color}
\usepackage{morefloats}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{ colorlinks,
linkcolor=darkblue,
filecolor=darkgreen,
urlcolor=darkred,
citecolor=darkblue}

@@

\begin{document}
% We need a better definitely shorter title...
\title{Approximate Bayesian Computation in Large Scale Structure: Constraining the Galaxy-Halo Connection}
\begin{abstract}
\todo{This is not a paper about hod, let's not start with describing hod. I suggest that we talk about parameter estimation in large scale structure in general and then move on to hod. Constraining hod is a just the first step in testing our methodology.} The Halo Occupation Distribution (HOD) model is a powerful prescription of how galaxies
populate dark matter halos that has had much success in reproducing large scale galaxy
clustering and understanding galaxy evolution within the context of the halo model.
Constraints on the parameters that describe the HOD model have so far been inferred using 
the standard practice of Bayesian inference with likelihood estimation that assumes a
functional Gaussian form of the likelihood.\todo{(Do we want to include analytic models for clustering statistics here? Yes, also we should say we use summary statistics that are not independent: clustering + group. Also there is no analytic model for the abundance of galaxy group richnesses)}  % I think we should keep it simple. There's no need to go into detail about everything in the abstract. This is already a pretty sizable abstract. 
Likelihood free inference methods such as the
Approximate Bayesian Computation (ABC) makes Bayesian inference possible without ever 
having to calculate the likelihood.

@@

Cosmology was revolutionized in the 1990s with the introduction of likelihoods---%
pro\-ba\-bil\-ities for the data given the theoretical model and parameter settings---%
for combining data from different surveys and performing principled inferences of
the cosmological parameters (CITE CITE).
Nowhere has this been more true than in cosmic microwave background studies,
where it is nearly possible to analytically evaluate a likelihood function that
involves no (or minimal) approximations (CITE Wandelt and others).

Fundamentally, the tractability of likelihood functions in cosmology flows from
the fact that the initial conditions are exceedingly close to Gaussian in form,
and many sources of measurement noise are also Gaussian. (may as well cite evidence for this too)
Likelihood functions are easier to write down and evaluate when things are closer to Gaussian,
so they get easier as the length scales get larger and the structure formation gets
younger.
Hence the suitability of CMB data for likelihood analyses.

@@

\section{Experiments}\label{sec:data}
\subsection{Mock observations}
\subsection{}
In order to test the va
%Plots showing 2pcf, gmf, ... of the mock data.
%How we compute the covariance matrix from mocks...
%Talk about error bars. Talk about poisson errors in group %populations .
%--------------------------------------------------------
% Covariance matrix of xi  
%--------------------------------------------------------
\begin{figure}
\includegraphics[width=0.5\textwidth]{xi_covariance_Mr20_Nmock500.png}
\caption{The reduce covariance matrix $C_{i,j}/\sqrt{C_{i,i} C_{j,j}}$ of 
the two-point correlation function, $\xi(r)$, estimated from 500 Halotools 
realizations. We describe the calculation of $\xi(r)$ in Section \todo{reference section}. \todo{binning and range of $r$}.}
\label{fig:2pcf_cov}
\end{figure}
%--------------------------------------------------------
% Covariance matrix of GMF 
%--------------------------------------------------------
\begin{figure}
  \includegraphics[width=0.5\textwidth]{gmf_covariance_Mr20_Nmock500.png}
\caption{The reduce covariance matrix $C_{i,j}/\sqrt{C_{i,i} C_{j,j}}$ of 
the Galaxy Multiplicity Function (GMF) estimated from 500 Halotools 
realizations. \todo{reference to GMF calculation in text}. 
\todo{binning and range of $r$}.}
\label{fig:gmf_cov}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%% SECTION: %%%%%%%%%%%%%%%%%%%%%%%
% SECTION: RESULTS 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}\label{sec:results}
Show the posteriors...
Show evolution of the posterior errorbars with iteration. Show epsilon. Convice that we are converging ...
Draw samples from the posterior, populate the mock catalogs with those samples, and compute the summary statistcs with them 
and demonstrate that they match the data...

\subsection{Comparison to MCMC}
\todo{Apples to apples comparison between ABC and MCMC.}
ABC-PMC using nbar and xi vs MCMC
%--------------------------------------------------------
% ABC-PMC Posterior likelihoods using nbar and xi
%--------------------------------------------------------
\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{nbar_xi_Mr20_Niter10000_Nburn100_mcmc_samples_test.png}
\caption{Posterior likelihood of HOD parameters $\log M_0$, 
$\sigma_{\log M}$, $\log M_{min}$, $\alpha$, and $\log M_1$ using the
standard Markov Chain Monte Carlo method as described in Section \todo{CITECITE}. 
\todo{THIS IS A SPACE MARKER NOT THE FINAL FIGURE.}}
\label{fig:post_abc_nbarxi}
\end{center}
\end{figure*}
%--------------------------------------------------------
% MC-MC Posterior likelihoods using nbar and xi
%--------------------------------------------------------
\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{nbar_xi_Mr20_now_t11.png}
\caption{Posterior likelihood of HOD parameters $\log M_0$, 
$\sigma_{\log M}$, $\log M_{min}$, $\alpha$, and $\log M_1$ using ABC-PMC 
with distance criteria defined by the observables $\bar{n}$ and $\xi(r)$.
\todo{This is the 12th iteration. NOT THE FINAL.}}
\label{fig:post_abc_nbarxi}
\end{center}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%% SECTION: %%%%%%%%%%%%%%%%%%%%%%%
% ABC INCLUDING GMF
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{ABC Including GMF}
ABC-PMC results using observables beyond galaxy clustering and nbar(z) GMF
%--------------------------------------------------------
% ABC-PMC Posterior likelihoods using nbar, xi and GMF 
%--------------------------------------------------------
\todo{Insert ABC-PMC Posterior Likelihoods using $\bar{n}$, $\xi(r)$ and GMF 
once it finishes running}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABC-PMC Posterior likelihoods using nbar and GMF 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{nbar_gmf5_Mr20_now_t18.png}
\caption{Posterior likelihood of HOD parameters $\log M_0$, 
$\sigma_{\log M}$, $\log M_{min}$, $\alpha$, and $\log M_1$ using ABC-PMC 
with distance criteria defined by the observables $\bar{n}$ and GMF.}
\label{fig:post_abc_nbargmf}
\end{center}
\end{figure*}
\begin{figure*}
\begin{center}
  \includegraphics[width=1.\textwidth]{tpcf_sigma.pdf}
\caption{1-$\sigma$(dark-shaded) and two-$\sigma$ ABC posterior predictions
for the two-point correlation function of galaxies. The black lines show the data-points.}
\label{fig:gmf_cov}
\end{center}
\end{figure*}

@@

\section*{Acknowledgments}

We would like to thank Mike Blanton, Boris Leidstadt, Micheal Schneider, 
Uros Seljak, Ross Fadely, Jeremy Tinker, Alex Malz, Patrick McDonald, Dan 
Foreman-Mackey, for productive and insightful discussions.

MJV was supported by NSF grant AST-1517237. CHH was supported by James 
Arthur Graduate Award Grant. DWH was supported by NSF (grants IIS-1124794 
and AST-1517237), NASA (grant NNX12AI50G), and the Moore-Sloan Data Science 
Environment at NYU.

\todo{We need to acknowledge AstroHackWeek.}
